---
source: Rmd
title: 'Cell Type Annotation'
teaching: 40
exercises: 20
---

:::::::::::::::::::::::::::::::::::::: questions

- How do we systematically find the genes that define each cluster?
- How do we assign cell type labels using known marker genes?
- How does automated annotation with SingleR work?
- When do manual and automated annotations disagree and what should we do?
- What are best practices for confident, reproducible annotation?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Run FindAllMarkers to identify differentially expressed genes for every cluster
- Interpret marker gene statistics: avg_log2FC, pct.1, pct.2, and p_val_adj
- Manually annotate clusters using known PBMC marker genes
- Use SingleR with the Monaco Immune reference for automated cell type annotation
- Compare manual and automated annotations and resolve discrepancies

::::::::::::::::::::::::::::::::::::::::::::::::

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(Seurat)
library(ggplot2)
library(patchwork)
library(SingleR)
library(celldex)
```

## Loading the Clustered Data

We start from the clustered Seurat object saved at the end of the previous
episode. Make sure the active identity is set to the resolution 0.5 clusters.

```{r load-data, eval=FALSE}
pbmc <- readRDS("pbmc_clustered.rds")
Idents(pbmc) <- "RNA_snn_res.0.5"
DimPlot(pbmc, reduction = "umap", label = TRUE, label.size = 5) + NoLegend()
```

:::::::::::::::::::::::::::::::::::::::::: spoiler

## Expected output

You should see the UMAP with approximately 9--11 clusters labeled by number,
matching the clustering results from the previous episode.

::::::::::::::::::::::::::::::::::::::::::::::::::


## Finding Marker Genes

Up to now, our clusters are just numbers. To give them biological meaning, we
need to find the genes that distinguish each cluster from all other clusters.
These are called **marker genes** -- genes that are significantly upregulated
in one cluster compared to the rest.

`FindAllMarkers()` runs a differential expression test for every cluster,
comparing cells in that cluster against all other cells:

```{r find-markers, eval=FALSE}
pbmc.markers <- FindAllMarkers(pbmc,
                               only.pos = TRUE,
                               min.pct = 0.25,
                               logfc.threshold = 0.25)
```

| Parameter | Value | Description |
|-----------|-------|-------------|
| `only.pos` | `TRUE` | Only return genes that are **upregulated** in the cluster (positive log fold change). We are interested in markers that define a cluster, not genes that are absent from it. |
| `min.pct` | `0.25` | Only test genes detected in at least 25% of cells in either the cluster or the rest. This speeds up the computation by skipping very rare genes that cannot be reliable markers. |
| `logfc.threshold` | `0.25` | Only test genes with at least a 0.25 log2 fold change between the cluster and the rest. This pre-filters out genes with trivially small differences. |

This step may take 1--2 minutes. The result is a data frame with one row per
gene per cluster. Let's look at the key columns:

```{r show-markers, eval=FALSE}
head(pbmc.markers)
```

:::::::::::::::::::::::::::::::::::::::::: spoiler

## Expected output

```output
               p_val avg_log2FC pct.1 pct.2 p_val_adj cluster gene
S100A9  0.000e+00      3.862 0.975 0.108 0.000e+00       1 S100A9
S100A8  0.000e+00      3.796 0.960 0.094 0.000e+00       1 S100A8
LYZ     0.000e+00      3.210 0.988 0.255 0.000e+00       1 LYZ
FCN1    0.000e+00      2.543 0.791 0.045 0.000e+00       1 FCN1
TYROBP  0.000e+00      2.372 0.995 0.320 0.000e+00       1 TYROBP
CD14    0.000e+00      2.123 0.683 0.034 0.000e+00       1 CD14
```

::::::::::::::::::::::::::::::::::::::::::::::::::

The columns mean:

| Column | Meaning |
|--------|:--------|
| `p_val` | Raw p-value from the Wilcoxon rank-sum test (default test). |
| `avg_log2FC` | Average log2 fold change between the cluster and all other cells. A value of 1.0 means the gene is on average 2x higher in this cluster. |
| `pct.1` | Fraction of cells **in the cluster** where the gene is detected. |
| `pct.2` | Fraction of cells **outside the cluster** where the gene is detected. |
| `p_val_adj` | Bonferroni-adjusted p-value. Use this for significance filtering (typically < 0.05). |
| `cluster` | Which cluster this gene is a marker for. |
| `gene` | The gene symbol. |

A good marker gene has a high `avg_log2FC`, high `pct.1` (expressed in most
cells of the cluster), low `pct.2` (not expressed in other clusters), and a
significant `p_val_adj`.

### Top markers per cluster

Let's extract the top 5 markers per cluster ranked by fold change:

```{r top-markers, eval=FALSE}
top5 <- pbmc.markers %>%
    group_by(cluster) %>%
    slice_max(n = 5, order_by = avg_log2FC)
top5
```

### Visualizing markers with a heatmap

A heatmap of top marker genes across all clusters shows the expression patterns
at a glance. Each column is a cell (grouped by cluster), each row is a gene,
and color intensity represents scaled expression:

```{r heatmap, eval=FALSE, fig.alt="Heatmap showing expression of the top 3 marker genes per cluster. Each cluster shows a distinct block of upregulated genes, confirming that clusters have distinct transcriptional identities."}
top3 <- pbmc.markers %>%
    group_by(cluster) %>%
    slice_max(n = 3, order_by = avg_log2FC)

DoHeatmap(pbmc, features = top3$gene) + NoLegend()
```

Each cluster should show a distinct block of highly expressed genes (bright
yellow/white) that are low in other clusters (dark purple). Clusters with very
similar heatmap profiles may represent subtypes of the same cell type.

### Dot plot summary

A dot plot provides a compact summary of marker expression across clusters.
The dot size encodes the percentage of cells expressing the gene, and the
color intensity encodes the average expression level:

```{r dotplot, eval=FALSE, fig.alt="Dot plot showing expression of top marker genes across all clusters. Each dot represents one gene in one cluster, with size indicating percent of cells expressing the gene and color indicating expression level."}
DotPlot(pbmc, features = unique(top3$gene)) + coord_flip() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


## Manual Annotation with Known Markers

Now we match the marker gene patterns to known PBMC cell types. The table below
lists the canonical markers for each expected cell type:

| Cell type | Key markers | Expected % |
|-----------|:-----------|:-----------|
| CD4+ T cells | CD3D, IL7R, CCR7 | ~35% |
| CD8+ T cells | CD3D, CD8A, CD8B | ~10% |
| B cells | MS4A1, CD79A | ~10% |
| CD14+ Monocytes | LYZ, CD14, S100A9 | ~20% |
| FCGR3A+ Monocytes | FCGR3A, MS4A7 | ~5% |
| NK cells | GNLY, NKG7, KLRD1 | ~10% |
| Dendritic cells | FCER1A, CST3 | ~3% |
| Platelets | PPBP, PF4 | ~2% |

Let's visualize these markers on the UMAP to see which clusters they map to:

```{r feature-markers, eval=FALSE, fig.alt="Grid of UMAP plots showing expression of eight canonical PBMC markers. Each marker lights up in a specific region of the UMAP corresponding to its cell type."}
FeaturePlot(pbmc,
            features = c("CD3D", "IL7R", "CD8A", "MS4A1",
                         "LYZ", "FCGR3A", "GNLY", "FCER1A"),
            ncol = 4)
```

And as violin plots to see expression across clusters:

```{r vln-markers, eval=FALSE, fig.alt="Violin plots showing expression of key PBMC markers across all clusters. Each marker shows high expression in one or two clusters corresponding to its known cell type."}
VlnPlot(pbmc,
        features = c("CD3D", "IL7R", "CD8A", "MS4A1",
                     "LYZ", "FCGR3A", "GNLY", "PPBP"),
        ncol = 4,
        pt.size = 0)
```

Based on these plots and the `FindAllMarkers` results, we can build a mapping
from cluster numbers to cell type names. Your exact cluster numbers will depend
on the random seed, but the logic is the same: find which clusters express each
set of markers.

```{r annotate, eval=FALSE}
# Build the cluster-to-cell-type mapping
# Adjust cluster IDs to match YOUR results
new.cluster.ids <- c(
    "CD4 T",           # 0: CD3D+, IL7R+
    "CD14 Mono",       # 1: LYZ+, CD14+, S100A9+
    "CD4 T",           # 2: CD3D+, IL7R+ (memory subset)
    "B",               # 3: MS4A1+, CD79A+
    "CD8 T",           # 4: CD3D+, CD8A+
    "NK",              # 5: GNLY+, NKG7+
    "FCGR3A Mono",     # 6: FCGR3A+, MS4A7+
    "DC",              # 7: FCER1A+, CST3+
    "Platelet",        # 8: PPBP+, PF4+
    "CD4 T"            # 9: CD3D+, IL7R+ (if present)
)
names(new.cluster.ids) <- levels(pbmc)
pbmc <- RenameIdents(pbmc, new.cluster.ids)
```

:::::::::::::::::::::::::::::::::::::::::: callout

## Annotation is iterative

Your first-pass annotation will not always be correct. This is normal. After
assigning labels, go back and check:

- Do the labels make biological sense? Is the proportion of each cell type
  reasonable for PBMCs?
- Are there clusters where multiple marker sets overlap? This may indicate
  a mixed cluster that should be sub-clustered at higher resolution.
- Are there clusters with no clear markers? These may be low-quality cells
  that slipped through QC, or a rare cell type you didn't expect.

Annotation is an iterative process: assign labels, verify with markers,
revise, and repeat until you are confident in the assignments.

:::::::::::::::::::::::::::::::::::::::::::::::::::

Visualize the final annotated UMAP:

```{r annotated-umap, eval=FALSE, fig.alt="UMAP plot with cells colored and labeled by cell type name. Distinct groups are labeled as CD4 T, CD8 T, B, CD14 Mono, FCGR3A Mono, NK, DC, and Platelet."}
DimPlot(pbmc, reduction = "umap", label = TRUE, repel = TRUE) + NoLegend()
```

Store the annotation in the metadata so it persists after saving:

```{r store-labels, eval=FALSE}
pbmc$manual_annotation <- Idents(pbmc)
```


## Automated Annotation with SingleR

Manual annotation requires expert knowledge of marker genes for every expected
cell type. **SingleR** offers an automated alternative: it compares the
expression profile of each cell to a labeled reference dataset and assigns the
label of the most similar reference sample.

### Download the reference

We use the **Monaco Immune Data** reference from the `celldex` package. This
reference contains bulk RNA-seq profiles of sorted human immune cell
populations, making it well-suited for PBMC annotation.

```{r get-reference, eval=FALSE}
ref <- celldex::MonacoImmuneData()
ref
```

:::::::::::::::::::::::::::::::::::::::::: spoiler

## Expected output

```output
class: SummarizedExperiment
dim: 46077 114
metadata(0):
assays(1): logcounts
rownames(46077): A1BG A1BG-AS1 ... ZZEF1 ZZZ3
rowData names(0):
colnames(114): DZQV_CD14+_monocytes DZQV_Exhausted_B_cells ...
  DZQV_T_regulatory_cells DZQV_Terminal_effector_CD8_T_cells
colData names(3): label.main label.fine label.ont
```

The reference has 114 samples across multiple immune cell types. The
`label.main` column has broad cell type labels (e.g., "Monocytes", "CD4+ T
cells") that match well with our expected PBMC types.

::::::::::::::::::::::::::::::::::::::::::::::::::

### Run SingleR

SingleR works on a `SingleCellExperiment` object. We convert our Seurat object
and then run the annotation:

```{r run-singler, eval=FALSE}
# Convert to SingleCellExperiment
sce <- as.SingleCellExperiment(DietSeurat(pbmc))

# Run SingleR with Monaco Immune reference
results <- SingleR(test = sce,
                   ref = ref,
                   labels = ref$label.main)
```

| Parameter | Value | Description |
|-----------|-------|-------------|
| `test` | `sce` | The query dataset as a SingleCellExperiment. SingleR uses the log-normalized expression values. |
| `ref` | `ref` | The reference dataset with known labels. |
| `labels` | `ref$label.main` | Which column of the reference metadata to use as cell type labels. `label.main` gives broad categories; `label.fine` gives more specific subtypes. |

This runs for 1--2 minutes. SingleR assigns a label to **every individual
cell**, not to clusters. Let's transfer the labels to our Seurat object:

```{r transfer-labels, eval=FALSE}
pbmc$singler_labels <- results$labels
```

Visualize the SingleR annotations on the UMAP:

```{r singler-umap, eval=FALSE, fig.alt="UMAP plot with cells colored by SingleR automated annotation labels from the Monaco Immune reference. The labels largely agree with the manual annotation, showing T cells, monocytes, B cells, NK cells, and other immune types."}
DimPlot(pbmc, group.by = "singler_labels", reduction = "umap",
        label = TRUE, repel = TRUE) + NoLegend()
```

### Comparing manual and automated annotations

The real power of annotation comes from comparing multiple approaches. Let's
create a cross-tabulation of our manual labels versus SingleR labels:

```{r compare, eval=FALSE}
table(Manual = pbmc$manual_annotation, SingleR = pbmc$singler_labels)
```

:::::::::::::::::::::::::::::::::::::::::: spoiler

## Expected output

You should see a table where most cells along the diagonal agree. For example,
cells you labeled "CD14 Mono" should predominantly be called "Monocytes" by
SingleR. Typical areas of agreement and disagreement:

- **Strong agreement**: CD14+ Monocytes, B cells, NK cells -- these have
  distinctive expression profiles and are called consistently by both methods.
- **Partial agreement**: CD4+ T cells may be split by SingleR into "CD4+ T
  cells" and "T regulatory cells"; CD8+ T cells may be split into subtypes.
  These finer distinctions from SingleR are often correct.
- **Possible disagreement**: FCGR3A+ Monocytes may be labeled as "Non-classical
  monocytes" by SingleR (which is the correct immunological term) or
  occasionally as "NK cells" if FCGR3A expression is high. Dendritic cells may
  also show disagreement if the cluster is small.

::::::::::::::::::::::::::::::::::::::::::::::::::

Visualize both annotations side by side:

```{r side-by-side, eval=FALSE, fig.alt="Two UMAP plots side by side comparing manual and SingleR annotations. Most clusters receive consistent labels between the two approaches."}
p1 <- DimPlot(pbmc, group.by = "manual_annotation", label = TRUE, repel = TRUE) +
    NoLegend() + ggtitle("Manual annotation")
p2 <- DimPlot(pbmc, group.by = "singler_labels", label = TRUE, repel = TRUE) +
    NoLegend() + ggtitle("SingleR annotation")
p1 + p2
```

When the two methods **agree**, you can be very confident in the label. When
they **disagree**, check the marker gene expression for the disputed cluster
and decide which label fits better. In general:

- Trust **manual annotation** when the markers are clear and well-established
  for the tissue type you are studying
- Trust **SingleR** when you are unfamiliar with the tissue or when the
  reference provides finer subtypes than you would identify manually

Save the final annotated object:

```{r save, eval=FALSE}
saveRDS(pbmc, file = "pbmc_annotated.rds")
```


## Annotation Best Practices

:::::::::::::::::::::::::::::::::::::::::: callout

## Guidelines for confident annotation

1. **Use multiple references.** No single reference is perfect. The `celldex`
   package provides several options for human immune cells:
   - `MonacoImmuneData()` -- sorted immune populations, best for PBMCs
   - `HumanPrimaryCellAtlasData()` (HPCA) -- broader coverage including
     non-immune cell types, but less specific for immune subtypes
   - `DatabaseImmuneCellExpressionData()` (DICE) -- another immune-focused
     reference with different experimental conditions

   Running SingleR with multiple references and checking for agreement
   increases confidence.

2. **Cross-validate automated labels with marker genes.** Never accept
   automated labels blindly. Always check that the assigned label is consistent
   with the expression of established markers for that cell type.

3. **Document your decisions.** Record which markers you used, which reference
   datasets, and any ambiguous cases you resolved manually. This makes your
   analysis reproducible and allows others to evaluate your annotation choices.

4. **Sub-cluster when needed.** If a cluster shows mixed marker expression
   (e.g., both CD4 and CD8 markers), it may contain two cell types that were
   merged at the current resolution. Re-run `FindClusters()` at higher
   resolution on just that subset of cells to resolve the mixture:

   ```r
   subset_cells <- subset(pbmc, idents = "mixed_cluster")
   subset_cells <- FindNeighbors(subset_cells, dims = 1:15)
   subset_cells <- FindClusters(subset_cells, resolution = 0.5)
   ```

:::::::::::::::::::::::::::::::::::::::::::::::::::


::::::::::::::::::::::::::::::::::::: challenge

## Challenge 1: Identifying a T Cell Subtype

One of the T cell clusters expresses **CD3D** and **SELL** (CD62L) at high
levels, but **CD69** expression is low. Use what you know about T cell biology
to determine what T cell subtype this is.

Check the expression of these additional markers in the cluster:

```{r challenge1, eval=FALSE}
VlnPlot(pbmc, features = c("CD3D", "SELL", "CCR7", "CD69", "IL7R"), ncol = 5, pt.size = 0)
```

:::::::::::::::::::::::: solution

This cluster represents **naive T cells**. The key evidence:

| Marker | Expression | Interpretation |
|--------|:----------|:---------------|
| CD3D | High | Confirms T cell identity |
| SELL (CD62L) | High | Expressed on naive and central memory T cells; required for lymph node homing |
| CCR7 | High | Another lymph node homing receptor; high in naive T cells |
| CD69 | Low | An early activation marker; low expression confirms these are NOT recently activated |
| IL7R (CD127) | High | IL-7 receptor, expressed on naive and memory T cells for homeostatic survival |

The combination of **SELL-high, CCR7-high, CD69-low** is the hallmark of
**naive T cells** that have not yet encountered their antigen. They recirculate
between blood and lymph nodes (hence the homing receptors) but have not been
activated (hence low CD69).

By contrast, **effector memory T cells** would be SELL-low, CCR7-low (they
patrol peripheral tissues rather than homing to lymph nodes) and may have
higher CD69 if recently activated.

This illustrates why multiple markers are needed: CD3D alone tells you it is
a T cell, but the combination of homing and activation markers tells you the
functional subtype.

:::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: challenge

## Challenge 2: Comparing Two SingleR References

Run SingleR with both `MonacoImmuneData` and `HumanPrimaryCellAtlasData`.
For which clusters do the two references disagree? Why might they disagree?

```{r challenge2, eval=FALSE}
# Monaco Immune reference (already computed above)
ref_monaco <- celldex::MonacoImmuneData()
results_monaco <- SingleR(test = sce, ref = ref_monaco, labels = ref_monaco$label.main)

# Human Primary Cell Atlas reference
ref_hpca <- celldex::HumanPrimaryCellAtlasData()
results_hpca <- SingleR(test = sce, ref = ref_hpca, labels = ref_hpca$label.main)

# Transfer labels
pbmc$singler_monaco <- results_monaco$labels
pbmc$singler_hpca <- results_hpca$labels

# Compare
table(Monaco = pbmc$singler_monaco, HPCA = pbmc$singler_hpca)
```

:::::::::::::::::::::::: solution

You should find that the two references **agree** on the major cell types
(T cells, B cells, monocytes, NK cells) but **disagree** on some clusters.
Common disagreements include:

**1. HPCA labels some immune cells with non-immune terms.** The HPCA reference
includes tissue types like "Tissue_stem_cells", "Epithelial_cells", and
"Smooth_muscle_cells" that are not present in a PBMC sample. When a cell does
not closely match any immune profile in the HPCA reference, it may be
incorrectly assigned to one of these non-immune categories. The Monaco
reference, which contains only immune cell types, avoids this problem.

**2. T cell subtype resolution differs.** Monaco distinguishes "CD4+ T cells"
from "CD8+ T cells" and further subtypes. HPCA may label all T cells
generically as "T_cells" without subtype distinction, because its T cell
reference profiles are less granular.

**3. Monocyte and dendritic cell boundaries differ.** The boundary between
monocytes and dendritic cells is biologically continuous, and different
references draw the line in slightly different places. Monocyte-derived DCs
may be called "Monocytes" by one reference and "DC" by the other.

**Lesson:** Use a reference that matches your tissue type. For PBMCs, the
Monaco Immune reference is more appropriate because it was built from sorted
immune populations. HPCA is better when you have a mixed tissue sample that
may contain both immune and non-immune cells.

:::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: keypoints

- FindAllMarkers identifies genes upregulated in each cluster versus all others; key output columns are avg_log2FC, pct.1, pct.2, and p_val_adj
- Manual annotation maps cluster marker genes to known cell-type-specific markers and requires domain expertise
- SingleR automates annotation by comparing each cell's expression profile to a labeled reference dataset
- Combining manual and automated approaches increases confidence; agreement between methods strongly supports the assigned label
- Annotation is iterative and should be cross-validated with multiple references and marker gene visualization

::::::::::::::::::::::::::::::::::::::::::::::::
