---
source: Rmd
title: 'Differential Expression Analysis'
teaching: 40
exercises: 20
---

:::::::::::::::::::::::::::::::::::::: questions

- What are the main approaches for differential expression testing in scRNA-seq?
- Why is pseudobulk analysis preferred over per-cell testing for multi-sample experiments?
- How do we create volcano plots and other visualizations to interpret DE results?
- How do we connect lists of DE genes to biological pathways using GO enrichment?
- How should we export our results for sharing and reproducibility?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Compare Wilcoxon, MAST, and pseudobulk DE testing approaches and choose the right one for your design
- Run pseudobulk differential expression using AggregateExpression
- Create volcano plots, feature plots, and violin plots to visualize DE results
- Perform gene ontology enrichment analysis with clusterProfiler and interpret the results
- Export Seurat objects, count matrices, and metadata for reproducibility and sharing

::::::::::::::::::::::::::::::::::::::::::::::::

## Setup

```{r setup, message=FALSE, warning=FALSE, eval=FALSE}
library(Seurat)
library(ggplot2)
library(patchwork)
library(dplyr)
library(clusterProfiler)
library(org.Hs.eg.db)
```

## Loading the Integrated Data

We continue from the integrated and annotated IFNB dataset produced in the
previous episode. Make sure the cell type annotation is set as the active
identity.

Set up the working directory to match the previous episodes:

```{r set-paths, eval=FALSE}
work_dir <- paste0(
    "/scratch/negishi/", Sys.getenv("USER"),
    "/scrna_workshop/"
)
setwd(work_dir)
```

```{r load-data, eval=FALSE}
ifnb <- readRDS(paste0(work_dir, "ifnb_annotated.rds"))
Idents(ifnb) <- "celltype"
DimPlot(ifnb, reduction = "umap", label = TRUE, repel = TRUE, split.by = "stim") +
    NoLegend()
```

:::::::::::::::::::::::::::::::::::::::::: spoiler

## Expected output

You should see two UMAP panels (CTRL and STIM) with the same cell type
structure in both. Cell types are labeled and colored consistently across
panels.

::::::::::::::::::::::::::::::::::::::::::::::::::


## DE Testing Methods

Seurat supports several statistical tests for differential expression. The
choice of method depends on your experimental design and whether you have
biological replicates.

| Method | Seurat parameter | Statistical model | Best for |
|--------|:----------------|:-----------------|:---------|
| **Wilcoxon rank-sum** | `test.use = "wilcox"` | Non-parametric rank test | Fast exploration, comparing clusters within a single sample |
| **MAST** | `test.use = "MAST"` | Hurdle model (logistic + Gaussian) | Accounts for dropout and cellular detection rate; per-cell testing with better modeling |
| **DESeq2 (pseudobulk)** | `test.use = "DESeq2"` | Negative binomial GLM on aggregated counts | Publication-quality results when biological replicates are available |

Let's start with the **Wilcoxon test** to quickly identify DE genes between
stimulated and control CD14+ monocytes:

```{r wilcox-de, eval=FALSE}
# Subset to CD14+ monocytes
mono <- subset(ifnb, idents = "CD14 Mono")

# Set condition as active identity
Idents(mono) <- "stim"

# Run Wilcoxon DE test
mono.wilcox <- FindMarkers(mono,
                           ident.1 = "STIM",
                           ident.2 = "CTRL",
                           test.use = "wilcox",
                           min.pct = 0.25,
                           logfc.threshold = 0.25)
head(mono.wilcox, 10)
```

| Parameter | Value | Description |
|-----------|-------|-------------|
| `ident.1` | `"STIM"` | The first group (numerator). Positive `avg_log2FC` means higher in STIM. |
| `ident.2` | `"CTRL"` | The second group (denominator). |
| `test.use` | `"wilcox"` | Wilcoxon rank-sum test. Fast, non-parametric, works well for exploration. |
| `min.pct` | `0.25` | Only test genes detected in at least 25% of cells in either group. |
| `logfc.threshold` | `0.25` | Only test genes with at least 0.25 log2FC between groups. |

:::::::::::::::::::::::::::::::::::::::::: spoiler

## Expected output

```output
              p_val avg_log2FC pct.1 pct.2    p_val_adj
IFIT1  0.000e+00      4.528 0.992 0.056 0.000e+00
ISG15  0.000e+00      4.345 0.995 0.186 0.000e+00
IFI6   0.000e+00      4.123 0.968 0.042 0.000e+00
IFIT3  0.000e+00      3.752 0.959 0.048 0.000e+00
IFI44L 0.000e+00      3.697 0.888 0.022 0.000e+00
IFIT2  0.000e+00      3.583 0.919 0.027 0.000e+00
RSAD2  0.000e+00      3.443 0.828 0.013 0.000e+00
ISG20  0.000e+00      3.159 0.940 0.124 0.000e+00
LY6E   0.000e+00      3.243 0.919 0.077 0.000e+00
MX1    0.000e+00      3.023 0.893 0.029 0.000e+00
```

The top genes are all interferon-stimulated genes (ISGs), as expected. But
notice that the p-values are extremely small (effectively zero). This is a
symptom of per-cell testing with thousands of cells.

::::::::::::::::::::::::::::::::::::::::::::::::::

```{r wilcox-count, eval=FALSE}
cat("Significant genes (padj < 0.05):", sum(mono.wilcox$p_val_adj < 0.05), "\n")
```

:::::::::::::::::::::::::::::::::::::::::: callout

## The pseudobulk revolution

The Wilcoxon and MAST tests treat **each cell as an independent observation**.
But cells from the same donor or the same sample are **not** independent --
they share the same genetic background, the same library preparation, and the
same sequencing run. This is called **pseudoreplication**, and it severely
inflates p-values.

For example, if you have 2,000 monocytes from one control sample and 2,000
from one stimulated sample, per-cell testing treats this as n = 4,000. But the
true sample size is n = 2 (one sample per condition). Every gene with even a
tiny difference will appear significant simply because of the enormous
"sample size."

**Pseudobulk analysis** solves this by aggregating all cells of the same type
from the same sample into a single expression profile. If you have 3 control
donors and 3 stimulated donors, you get 3 pseudobulk replicates per condition
-- the same as a bulk RNA-seq experiment with n = 3 per group. This gives
proper type I error control and realistic p-values.

**Always use pseudobulk** when you have biological replicates. Use per-cell
tests only for exploratory analysis or when you have a single sample per
condition and have no other option.

:::::::::::::::::::::::::::::::::::::::::::::::::::


## Pseudobulk Analysis

Pseudobulk aggregation sums the raw UMI counts for each gene across all cells
of the same cell type from the same sample, producing a bulk-like count matrix
with one column per sample-celltype combination.

```{r aggregate, eval=FALSE}
# Create pseudobulk counts
# Group by cell type, condition, and sample (orig.ident)
bulk <- AggregateExpression(ifnb,
                            group.by = c("celltype", "stim", "orig.ident"),
                            return.seurat = TRUE)
bulk
```

| Parameter | Value | Description |
|-----------|-------|-------------|
| `group.by` | `c("celltype", "stim", "orig.ident")` | Metadata columns to group cells by. Each unique combination of cell type, condition, and sample becomes one pseudobulk column. |
| `return.seurat` | `TRUE` | Return a Seurat object (instead of a raw matrix) so we can use `FindMarkers` on it. |

:::::::::::::::::::::::::::::::::::::::::: spoiler

## Expected output

```output
An object of class Seurat
14053 features across N samples within 1 assay
Active assay: RNA (14053 features, 0 variable features)
 1 layer present: counts
```

The number of samples (N) depends on the number of unique cell type + condition
+ donor combinations in the dataset.

::::::::::::::::::::::::::::::::::::::::::::::::::

Now we can run DESeq2-based DE testing on the pseudobulk object. We subset to
CD14+ monocytes and compare conditions:

```{r pseudobulk-de, eval=FALSE}
# Subset pseudobulk to CD14 Mono samples
bulk.mono <- subset(bulk, celltype == "CD14 Mono")

# Set condition as active identity
Idents(bulk.mono) <- "stim"

# Run DESeq2 on pseudobulk counts
mono.pseudo <- FindMarkers(bulk.mono,
                           ident.1 = "STIM",
                           ident.2 = "CTRL",
                           test.use = "DESeq2")
head(mono.pseudo, 10)
```

```{r pseudo-count, eval=FALSE}
cat("Significant genes (padj < 0.05):", sum(mono.pseudo$p_val_adj < 0.05, na.rm = TRUE), "\n")
```

:::::::::::::::::::::::::::::::::::::::::: spoiler

## Expected output

The pseudobulk analysis typically identifies **fewer significant genes** than
the per-cell Wilcoxon test (often hundreds instead of thousands). The top genes
are still ISGs (IFIT1, ISG15, IFI6), but the p-values are larger and more
realistic because the effective sample size is the number of donors, not the
number of cells.

::::::::::::::::::::::::::::::::::::::::::::::::::


## Visualization and Interpretation

### Volcano plot

A volcano plot shows the relationship between statistical significance
(-log10 p-value, y-axis) and biological effect size (log2 fold change,
x-axis). Each point is a gene. Genes in the upper corners are both
statistically significant and biologically meaningful.

```{r volcano, eval=FALSE, fig.alt="Volcano plot showing differential expression results for stimulated versus control CD14 monocytes. Significant upregulated genes appear in the upper right (including ISG15, IFIT1, IFI6), significant downregulated genes in the upper left, and non-significant genes cluster near the bottom center."}
# Use Wilcoxon results for visualization (more genes to plot)
mono.wilcox$gene <- rownames(mono.wilcox)
mono.wilcox$significance <- case_when(
    mono.wilcox$p_val_adj < 0.05 & mono.wilcox$avg_log2FC > 0.5  ~ "Up",
    mono.wilcox$p_val_adj < 0.05 & mono.wilcox$avg_log2FC < -0.5 ~ "Down",
    TRUE ~ "NS"
)

ggplot(mono.wilcox, aes(x = avg_log2FC, y = -log10(p_val_adj), color = significance)) +
    geom_point(size = 0.5, alpha = 0.6) +
    scale_color_manual(values = c("Up" = "firebrick", "Down" = "steelblue", "NS" = "grey70")) +
    geom_vline(xintercept = c(-0.5, 0.5), linetype = "dashed", color = "grey40") +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
    labs(x = "Log2 fold change (STIM vs CTRL)",
         y = "-log10(adjusted p-value)",
         title = "CD14+ Monocytes: STIM vs CTRL",
         color = "Direction") +
    theme_minimal()
```

How to read the volcano plot:

- **Upper right**: genes significantly upregulated in stimulated cells (ISGs)
- **Upper left**: genes significantly downregulated in stimulated cells
- **Bottom center**: non-significant genes (low fold change or high p-value)
- **Dashed lines**: significance cutoffs (|log2FC| > 0.5 and padj < 0.05)

### Feature and violin plots

Visualize specific DE genes on the UMAP and across conditions:

```{r de-viz, eval=FALSE, fig.alt="FeaturePlot showing ISG15 expression split by control and stimulated conditions. ISG15 is nearly absent in control cells but strongly expressed across multiple cell types in stimulated cells, especially monocytes."}
FeaturePlot(ifnb,
            features = "ISG15",
            split.by = "stim",
            cols = c("grey85", "firebrick"))
```

```{r de-violin, eval=FALSE, fig.alt="Violin plots showing ISG15, IFIT1, and MX1 expression across cell types, split by control and stimulated conditions. All three genes are dramatically upregulated in the stimulated condition across multiple cell types."}
VlnPlot(ifnb,
        features = c("ISG15", "IFIT1", "MX1"),
        split.by = "stim",
        pt.size = 0,
        ncol = 3)
```

### Dot plot across cell types

A dot plot can show the top DE genes across all cell types simultaneously,
revealing which responses are shared and which are cell-type-specific:

```{r de-dotplot, eval=FALSE, fig.alt="Dot plot showing expression of the top 10 DE genes across all cell types, split by condition. Interferon-stimulated genes are consistently upregulated across all cell types in the stimulated condition."}
top_de_genes <- head(rownames(mono.wilcox), 10)
DotPlot(ifnb,
        features = top_de_genes,
        split.by = "stim",
        cols = c("steelblue", "firebrick")) +
    coord_flip() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


## Functional Enrichment

A list of hundreds of DE genes is difficult to interpret gene by gene.
**Gene Ontology (GO) enrichment analysis** asks: are the DE genes enriched
for specific biological processes? This connects individual gene changes to
broader biological themes.

### Converting gene symbols to Entrez IDs

clusterProfiler requires Entrez gene IDs. We convert from gene symbols using
`bitr()`:

```{r convert-ids, eval=FALSE}
# Get significant upregulated genes
sig_genes <- mono.wilcox %>%
    filter(p_val_adj < 0.05, avg_log2FC > 0.5) %>%
    rownames()

cat("Number of significant upregulated genes:", length(sig_genes), "\n")

# Convert gene symbols to Entrez IDs
gene_ids <- bitr(sig_genes,
                 fromType = "SYMBOL",
                 toType = "ENTREZID",
                 OrgDb = org.Hs.eg.db)
head(gene_ids)
```

| Parameter | Value | Description |
|-----------|-------|-------------|
| `fromType` | `"SYMBOL"` | Input ID type (gene symbols like ISG15, IFIT1). |
| `toType` | `"ENTREZID"` | Output ID type (Entrez gene IDs like 9636, 3434). |
| `OrgDb` | `org.Hs.eg.db` | Organism annotation database (human). |

### Running GO enrichment

```{r enrichgo, eval=FALSE}
go_results <- enrichGO(gene = gene_ids$ENTREZID,
                       OrgDb = org.Hs.eg.db,
                       ont = "BP",
                       pAdjustMethod = "BH",
                       qvalueCutoff = 0.05,
                       readable = TRUE)
```

| Parameter | Value | Description |
|-----------|-------|-------------|
| `gene` | `gene_ids$ENTREZID` | Vector of Entrez gene IDs to test. |
| `OrgDb` | `org.Hs.eg.db` | Organism annotation database. |
| `ont` | `"BP"` | GO ontology category: "BP" (Biological Process), "MF" (Molecular Function), or "CC" (Cellular Component). |
| `pAdjustMethod` | `"BH"` | Benjamini-Hochberg method for multiple testing correction. |
| `qvalueCutoff` | `0.05` | Only return terms with q-value below this threshold. |
| `readable` | `TRUE` | Convert Entrez IDs in the output back to gene symbols for readability. |

Visualize the top enriched GO terms with a dot plot:

```{r go-dotplot, eval=FALSE, fig.alt="Dot plot showing the top 15 enriched GO biological process terms for upregulated genes in stimulated monocytes. The top terms include defense response to virus, type I interferon signaling pathway, response to virus, and innate immune response."}
dotplot(go_results, showCategory = 15) +
    ggtitle("GO Biological Process: STIM vs CTRL CD14+ Monocytes")
```

:::::::::::::::::::::::::::::::::::::::::: spoiler

## Expected output

The top enriched GO terms should include:

- **defense response to virus**
- **type I interferon signaling pathway**
- **response to virus**
- **innate immune response**
- **negative regulation of viral genome replication**
- **cytokine-mediated signaling pathway**
- **response to interferon-beta**

These terms are entirely consistent with IFN-beta stimulation. IFN-beta is a
type I interferon that activates antiviral defense programs, and the GO
enrichment confirms that the DE genes are concentrated in exactly these
pathways.

::::::::::::::::::::::::::::::::::::::::::::::::::

### KEGG pathway enrichment

For a complementary view, we can also test KEGG pathways:

```{r kegg, eval=FALSE}
kegg_results <- enrichKEGG(gene = gene_ids$ENTREZID,
                           organism = "hsa",
                           pAdjustMethod = "BH",
                           qvalueCutoff = 0.05)
head(kegg_results, 5)
```

:::::::::::::::::::::::::::::::::::::::::: spoiler

## Expected output

Top KEGG pathways should include entries like "Influenza A", "Hepatitis C",
"RIG-I-like receptor signaling pathway", and "NOD-like receptor signaling
pathway" -- all pathways involving the interferon/antiviral response.

::::::::::::::::::::::::::::::::::::::::::::::::::


## Exporting Results

At the end of an analysis, you should export your results in formats that
support reproducibility, sharing, and downstream use.

### Save the annotated Seurat object

```{r save-rds, eval=FALSE}
saveRDS(ifnb, file = paste0(work_dir, "ifnb_annotated.rds"))
```

### Export the count matrix and metadata

For the PBMC 10k dataset from earlier episodes, you can export the raw counts
and metadata as CSV files:

```{r export-csv, eval=FALSE}
# Export raw count matrix (warning: can be large for big datasets)
counts_matrix <- GetAssayData(pbmc, layer = "counts")
writeMM(counts_matrix, file = "counts.mtx")

# Export metadata
write.csv(pbmc@meta.data, file = "metadata.csv", quote = FALSE)

# Export gene names
write.csv(data.frame(gene = rownames(pbmc)), file = "features.csv",
          row.names = FALSE, quote = FALSE)
```

For smaller datasets, you can export as a dense CSV:

```{r export-dense, eval=FALSE}
write.csv(as.matrix(GetAssayData(pbmc, layer = "counts")),
          file = "counts_dense.csv")
```

### Format conversion for Python users

If collaborators use the Python/Scanpy ecosystem, you can convert your Seurat
object to AnnData format using the `sceasy` package:

```{r convert-anndata, eval=FALSE}
# Install sceasy if needed: remotes::install_github("cellgeni/sceasy")
library(sceasy)
sceasy::convertFormat(pbmc,
                      from = "seurat",
                      to = "anndata",
                      outFile = "pbmc_annotated.h5ad")
```

### GEO submission

For submitting data to NCBI GEO, you typically need:

- **Raw count matrix** (genes x cells) in MTX or CSV format
- **Cell metadata** (barcodes, cell types, conditions) as a TSV/CSV
- **Processing description** (software versions, parameters, filtering criteria)

:::::::::::::::::::::::::::::::::::::::::: callout

## Reproducibility checklist

Before sharing or publishing your analysis, ensure you have:

- **Saved your R session info** so others know exactly which package versions
  you used:

  ```r
  writeLines(capture.output(sessionInfo()), "session_info.txt")
  ```

- **Pinned package versions** in your analysis scripts or used `renv` to create
  a reproducible environment.

- **Saved intermediate RDS files** at each major step (`pbmc_filtered.rds`,
  `pbmc_normalized.rds`, `pbmc_clustered.rds`, `pbmc_annotated.rds`) so others
  can enter the pipeline at any point.

- **Documented all parameter choices**: QC thresholds, number of PCs,
  clustering resolution, normalization method, integration method, and DE test
  used.

- **Recorded the random seed** if you used one (e.g., `set.seed(42)` before
  UMAP or clustering).

:::::::::::::::::::::::::::::::::::::::::::::::::::


::::::::::::::::::::::::::::::::::::: challenge

## Challenge 1: Wilcoxon vs. Pseudobulk

Run both Wilcoxon (per-cell) and pseudobulk (DESeq2) DE testing for stimulated
vs. control CD14+ monocytes. Compare the number of significant genes and the
p-value distributions.

```{r challenge1, eval=FALSE}
# Wilcoxon (already computed above)
n_wilcox <- sum(mono.wilcox$p_val_adj < 0.05)
cat("Wilcoxon significant genes:", n_wilcox, "\n")

# Pseudobulk (already computed above)
n_pseudo <- sum(mono.pseudo$p_val_adj < 0.05, na.rm = TRUE)
cat("Pseudobulk significant genes:", n_pseudo, "\n")

# Compare p-value distributions for shared genes
shared_genes <- intersect(rownames(mono.wilcox), rownames(mono.pseudo))
comparison <- data.frame(
    gene = shared_genes,
    wilcox_pval = mono.wilcox[shared_genes, "p_val_adj"],
    pseudo_pval = mono.pseudo[shared_genes, "p_val_adj"]
)
comparison <- comparison[complete.cases(comparison), ]

ggplot(comparison, aes(x = -log10(wilcox_pval), y = -log10(pseudo_pval))) +
    geom_point(size = 0.5, alpha = 0.5) +
    geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
    labs(x = "-log10(p-value) Wilcoxon",
         y = "-log10(p-value) Pseudobulk",
         title = "P-value comparison: Wilcoxon vs Pseudobulk") +
    theme_minimal()
```

Which method gives more conservative results? Why?

:::::::::::::::::::::::: solution

You should find that:

- **Wilcoxon** identifies thousands of significant genes (often 2,000--4,000+)
- **Pseudobulk** identifies far fewer (often 200--800)

The scatter plot shows that most points fall **below** the diagonal (the red
dashed line), meaning pseudobulk p-values are consistently **larger** (less
significant) than Wilcoxon p-values for the same genes.

This is expected and correct. The Wilcoxon test has inflated significance
because it treats each cell as an independent observation. With ~2,000
monocytes per condition, even tiny expression differences become
"significant." The pseudobulk test treats each **donor** as one observation,
giving a realistic sample size and honest p-values.

The genes that are significant in **both** methods are the most trustworthy --
these are genes with large, robust effects that survive the more stringent
pseudobulk test. The genes significant only in Wilcoxon but not in pseudobulk
are likely false positives driven by pseudoreplication.

**Rule of thumb:** Use Wilcoxon for quick exploration and hypothesis
generation. Use pseudobulk for any result you plan to report or publish.

:::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: challenge

## Challenge 2: GO Enrichment of IFN-beta Response Genes

Run GO enrichment on the top 200 upregulated genes (by fold change) in
stimulated vs. control CD14+ monocytes. What are the top 5 biological
processes? Do they make sense for IFN-beta stimulation?

```{r challenge2, eval=FALSE}
# Get top 200 upregulated genes by fold change
top200 <- mono.wilcox %>%
    filter(p_val_adj < 0.05, avg_log2FC > 0) %>%
    arrange(desc(avg_log2FC)) %>%
    head(200)
top200_genes <- rownames(top200)

# Convert to Entrez IDs
top200_ids <- bitr(top200_genes,
                   fromType = "SYMBOL",
                   toType = "ENTREZID",
                   OrgDb = org.Hs.eg.db)

# Run GO enrichment
go_top200 <- enrichGO(gene = top200_ids$ENTREZID,
                      OrgDb = org.Hs.eg.db,
                      ont = "BP",
                      pAdjustMethod = "BH",
                      qvalueCutoff = 0.05,
                      readable = TRUE)

# Show top 5 terms
head(go_top200, 5)

# Visualize
dotplot(go_top200, showCategory = 10) +
    ggtitle("GO BP: Top 200 upregulated genes in STIM CD14+ Monocytes")
```

:::::::::::::::::::::::: solution

The top 5 biological processes should include terms like:

| Rank | GO Term | Interpretation |
|:----:|:--------|:--------------|
| 1 | defense response to virus | Direct antiviral effector programs activated by IFN-beta |
| 2 | type I interferon signaling pathway | The canonical pathway triggered by IFN-beta binding to IFNAR |
| 3 | response to virus | Broad viral response including both detection and effector mechanisms |
| 4 | negative regulation of viral genome replication | Specific antiviral effectors (MX1, IFIT proteins) that block viral replication |
| 5 | innate immune response | General innate immunity activation, of which IFN signaling is a central component |

These results make complete biological sense. IFN-beta (interferon beta) is a
**type I interferon** that signals through the IFNAR receptor complex,
activating the JAK-STAT pathway, which turns on hundreds of
interferon-stimulated genes (ISGs). These ISGs encode proteins that directly
inhibit viral replication (MX1, IFIT1), degrade viral RNA (ISG20, OAS
proteins), and alert neighboring cells to the viral threat (chemokines,
cytokines).

The GO enrichment confirms that our DE analysis is capturing the expected
biology: the top upregulated genes in stimulated monocytes are overwhelmingly
involved in antiviral and interferon signaling pathways. This serves as
important biological validation that the computational pipeline (integration,
DE testing, enrichment) is working correctly.

:::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: keypoints

- Per-cell DE tests (Wilcoxon, MAST) treat each cell as independent, inflating p-values through pseudoreplication
- Pseudobulk analysis aggregates cells by sample and cell type, providing statistically valid results with proper type I error control
- Volcano plots visualize the relationship between effect size (log2FC) and significance (-log10 p-value) for all tested genes
- Gene ontology enrichment connects lists of DE genes to biological pathways, providing functional interpretation of differential expression results
- Always export session info, save intermediate objects, and document parameter choices for reproducibility

::::::::::::::::::::::::::::::::::::::::::::::::
