---
source: Rmd
title: 'Dimensionality Reduction and Clustering'
teaching: 45
exercises: 15
---

:::::::::::::::::::::::::::::::::::::: questions

- Why do we need to reduce the dimensions of scRNA-seq data before clustering?
- How do we choose the right number of principal components?
- What does UMAP show us and what can it NOT tell us?
- How does graph-based clustering work and what does the resolution parameter control?
- How do we evaluate whether our clustering resolution is appropriate?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Run PCA on variable features and select an appropriate number of PCs using the elbow plot
- Generate a UMAP embedding for 2D visualization of cell relationships
- Construct a shared nearest neighbor graph and identify clusters with the Louvain algorithm
- Explore multiple clustering resolutions and use clustree to assess cluster stability
- Visualize known marker genes on the UMAP to preview cell type identity

::::::::::::::::::::::::::::::::::::::::::::::::

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(Seurat)
library(ggplot2)
library(patchwork)
library(clustree)
```

## Loading the Normalized Data

We start from the normalized, scaled Seurat object saved at the end of the
previous episode. This object has the `counts`, `data`, and `scale.data`
layers populated, with 2,000 variable features selected.

```{r load-data, eval=FALSE}
pbmc <- readRDS("pbmc_normalized.rds")
pbmc
```

:::::::::::::::::::::::::::::::::::::::::: spoiler

## Expected output

```output
An object of class Seurat
23694 features across 10000 samples within 1 assay
Active assay: RNA (23694 features, 2000 variable features)
 3 layers present: counts, data, scale.data
```

::::::::::::::::::::::::::::::::::::::::::::::::::


## Principal Component Analysis (PCA)

Our dataset has ~2,000 variable features, but many of these genes are
correlated -- they go up and down together across cells because they are
co-regulated in the same cell types. **PCA (Principal Component Analysis)**
exploits these correlations to compress the data into a smaller set of
**principal components (PCs)**. Each PC is a linear combination of genes that
captures a distinct axis of variation in the data.

The key properties of PCA:

- **PC1** captures the direction of greatest variance across all cells
- **PC2** captures the next greatest variance, orthogonal to PC1
- Each subsequent PC captures decreasing amounts of variance
- The first 10--20 PCs typically capture the major biological structure (cell
  types), while later PCs are dominated by noise

```{r run-pca, eval=FALSE}
pbmc <- RunPCA(pbmc, features = VariableFeatures(object = pbmc))
```

| Parameter | Value | Description |
|-----------|-------|-------------|
| `features` | `VariableFeatures(object = pbmc)` | Which genes to use for PCA. We use the 2,000 variable features selected in the previous episode. |

By default, Seurat computes 50 PCs. Let's examine which genes drive the first
two components:

```{r viz-loadings, eval=FALSE, fig.alt="Bar plots showing the top genes with the highest positive and negative loadings for PC1 and PC2. Genes like CST3 and LYZ appear in PC1, while CD3D and IL32 appear in PC2."}
VizDimLoadings(pbmc, dims = 1:2, reduction = "pca")
```

Each bar represents a gene's **loading** (contribution) to that PC. Genes with
large positive loadings push cells in one direction along the PC, while genes
with large negative loadings push them in the opposite direction. For PC1, you
should see monocyte markers (CST3, LYZ, S100A9) on one side and lymphocyte
markers on the other. This tells us that **PC1 separates myeloid from
lymphoid cells** -- the largest source of variation in PBMCs.

We can also visualize PCs as heatmaps to see which genes and cells contribute
most:

```{r dim-heatmap, eval=FALSE, fig.alt="Grid of nine heatmaps showing gene expression patterns for PC1 through PC9. Each heatmap shows the top genes ordered by loading with cells ordered by PC score, revealing distinct expression programs."}
DimHeatmap(pbmc, dims = 1:9, cells = 500, balanced = TRUE)
```

| Parameter | Value | Description |
|-----------|-------|-------------|
| `dims` | `1:9` | Which PCs to plot. |
| `cells` | `500` | Number of cells to include (randomly sampled from each extreme of the PC). |
| `balanced` | `TRUE` | Plot equal numbers of cells from the positive and negative ends of each PC. |

Each heatmap shows the top genes (rows) ordered by their loading, with cells
(columns) ordered by their PC score. Clear blocks of correlated expression
indicate PCs that capture real biological structure. PCs that look noisy
(no clear pattern) are capturing technical variation or random noise.

### Choosing the number of PCs

Not all 50 PCs are informative. Later PCs capture decreasing amounts of
variance and eventually become dominated by noise. We need to decide how many
PCs to retain for downstream steps (UMAP and clustering).

The **elbow plot** shows the standard deviation explained by each PC:

```{r elbow, eval=FALSE, fig.alt="Line plot of standard deviation versus principal component number. The curve drops steeply for the first 10 PCs then gradually flattens, with an elbow around PC 12 to 15."}
ElbowPlot(pbmc, ndims = 30)
```

Look for the **elbow** -- the point where the curve transitions from steep
decline to a flatter plateau. PCs before the elbow capture substantial
biological signal; PCs after the elbow are mostly noise.

:::::::::::::::::::::::::::::::::::::::::: callout

## How many PCs?

For the PBMC 10k dataset, the elbow typically falls around **PC 12--15**. We
will use **15 PCs** for the rest of this episode.

In practice, the exact number is not critical. Using 10 vs. 20 PCs usually
produces very similar clustering results for well-separated cell types.
The choice matters more for subtle distinctions between closely related
populations.

**When in doubt, err on the side of including more PCs** (e.g., 20). Using
too few PCs risks losing biological signal from rarer cell types. Using a few
extra PCs adds a small amount of noise but rarely changes the major structure.

An alternative to visual inspection is the **JackStraw** procedure
(`JackStrawPlot()`), which uses a permutation test to identify statistically
significant PCs. However, it is computationally expensive for large datasets
and is rarely necessary when the elbow is clear.

:::::::::::::::::::::::::::::::::::::::::::::::::::


## UMAP Visualization

PCA reduces our data from 2,000 dimensions to 15, but we still cannot
visualize 15 dimensions directly. **UMAP** (Uniform Manifold Approximation and
Projection) further reduces the data to just 2 dimensions for plotting. Unlike
PCA, UMAP is a **non-linear** method: it can unfold complex relationships that
PCA cannot capture, placing similar cells close together and dissimilar cells
far apart in 2D space.

```{r run-umap, eval=FALSE}
pbmc <- RunUMAP(pbmc, dims = 1:15)
```

| Parameter | Value | Description |
|-----------|-------|-------------|
| `dims` | `1:15` | Which PCs to use as input. We use the 15 PCs selected above. |

Visualize the UMAP embedding:

```{r umap-plot, eval=FALSE, fig.alt="UMAP plot showing approximately 10000 cells arranged in distinct clusters. Cells are colored by their identity, forming several well-separated groups."}
DimPlot(pbmc, reduction = "umap")
```

At this point, cells are not yet clustered (all are the same color). But you
can already see that cells organize into distinct groups on the UMAP. These
groups correspond to different cell types.

### UMAP parameters

UMAP has two key parameters that control the appearance of the plot:

| Parameter | Default | Effect |
|-----------|:--------|:-------|
| `n.neighbors` | 30 | Controls the balance between local and global structure. **Low values** (5--15) emphasize local neighborhoods, producing tighter clusters. **High values** (50--200) emphasize global structure, producing smoother layouts. |
| `min.dist` | 0.3 | Controls how tightly UMAP packs points. **Low values** (0.01--0.1) produce dense, compact clusters. **High values** (0.5--1.0) spread clusters out more evenly. |

The defaults work well for most datasets. Adjusting these parameters changes
the **visual appearance** of the plot but does **not** change the underlying
data or the clustering results (which operate on PCA space, not UMAP space).

:::::::::::::::::::::::::::::::::::::::::: callout

## UMAP interpretation pitfalls

UMAP plots are powerful visualization tools, but they can be misleading if
over-interpreted. Three things you should **NOT** infer from a UMAP:

1. **Distances between clusters do not reflect true biological similarity.**
   Two clusters that appear far apart on the UMAP may actually be
   transcriptionally very similar, and vice versa. UMAP distorts global
   distances to preserve local neighborhoods.

2. **Cluster size on the plot does not reflect population size.** A large,
   spread-out cluster may contain fewer cells than a small, compact one.
   UMAP adjusts spacing to show local structure, not proportional
   representation.

3. **Different random seeds produce different layouts.** Running UMAP twice
   with different seeds will give a different arrangement of clusters. The
   overall groupings will be the same (the same cells will cluster together),
   but their positions and orientations on the plot will differ. Never compare
   spatial positions across two separate UMAP runs.

**Always verify biological conclusions with quantitative methods** (marker
gene expression, differential expression tests, cluster composition
statistics) rather than relying on visual impressions from UMAP alone.

:::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::::: callout

## t-SNE as an alternative

Before UMAP became standard, **t-SNE** (t-distributed Stochastic Neighbor
Embedding) was the most common visualization method for scRNA-seq. You can run
it with `RunTSNE(pbmc, dims = 1:15)`. t-SNE is similar in purpose to UMAP but
tends to be slower and produces rounder, more separated clusters. The same
interpretation caveats apply. Most current analyses use UMAP because it is
faster, better preserves global structure, and scales better to large datasets.

:::::::::::::::::::::::::::::::::::::::::::::::::::


## Graph-Based Clustering

With the PCA embedding computed, we can now group cells into clusters. Seurat
uses a **graph-based clustering** approach that works in two steps:

### Step 1: Build a neighbor graph

`FindNeighbors()` constructs a **K-nearest neighbor (KNN) graph** in PCA
space. For each cell, it identifies its *k* most similar cells based on
Euclidean distance in the first 15 PCs. These connections are then refined
into a **shared nearest neighbor (SNN) graph**, where the connection strength
between two cells depends on how much their neighborhoods overlap.

```{r find-neighbors, eval=FALSE}
pbmc <- FindNeighbors(pbmc, dims = 1:15)
```

| Parameter | Value | Description |
|-----------|-------|-------------|
| `dims` | `1:15` | Which PCs to use for computing distances. Must match the PCs used for UMAP. |
| `k.param` | `20` (default) | Number of nearest neighbors per cell. Higher values produce a more connected graph with smoother clustering. Lower values preserve finer local structure. |

### Step 2: Detect communities

`FindClusters()` applies a **community detection algorithm** to the SNN graph.
The algorithm finds groups of cells that are densely connected to each other
but sparsely connected to the rest of the graph. The **resolution** parameter
controls the granularity of clustering:

```{r find-clusters, eval=FALSE}
pbmc <- FindClusters(pbmc, resolution = 0.5)
```

| Parameter | Value | Description |
|-----------|-------|-------------|
| `resolution` | `0.5` | Controls the number of clusters. Lower values (0.1--0.3) produce fewer, larger clusters. Higher values (1.0--2.0) produce more, smaller clusters. |
| `algorithm` | `1` (default) | Community detection method. `1` = Louvain (default, fast and reliable). `4` = Leiden (more robust against poorly connected clusters, requires the `leidenalg` Python package). |

:::::::::::::::::::::::::::::::::::::::::: spoiler

## Expected output

At resolution 0.5, you should get approximately **9--12 clusters**, numbered
starting from 0. The exact number may vary slightly depending on random seed
and software versions.

::::::::::::::::::::::::::::::::::::::::::::::::::

The cluster assignments are stored in the object metadata. Let's check them:

```{r show-clusters, eval=FALSE}
head(Idents(pbmc))
table(Idents(pbmc))
```

:::::::::::::::::::::::::::::::::::::::::: spoiler

## Expected output

```output
AAACCCAAGAAACACT-1 AAACCCAAGAAACCAT-1 AAACCCAAGAAACTGT-1 AAACCCAAGAAAGCGA-1
                 0                  2                  4                  0
AAACCCAAGAAAGTTG-1 AAACCCAAGAACAATC-1
                 4                  1

   0    1    2    3    4    5    6    7    8    9
3412 1897 1335 1022  858  547  391  277  152  109
```

Cluster 0 is the largest (T cells, which are the most abundant PBMC type).
Cluster 9 is the smallest (likely a rare population like platelets or
dendritic cells).

::::::::::::::::::::::::::::::::::::::::::::::::::

Now let's visualize the clusters on the UMAP:

```{r umap-clusters, eval=FALSE, fig.alt="UMAP plot with cells colored by cluster identity. Approximately 10 distinct colored groups are visible, each forming a separate region on the plot, with cluster numbers labeled at each group center."}
DimPlot(pbmc, reduction = "umap", label = TRUE, label.size = 5) +
    NoLegend()
```

### Exploring multiple resolutions

The resolution parameter is the most important choice in clustering. Too low
and distinct cell types are merged; too high and single cell types are split
into artificial sub-clusters. The best way to evaluate the right resolution is
to try several and compare.

```{r multi-resolution, eval=FALSE}
pbmc <- FindClusters(pbmc, resolution = 0.2)
pbmc <- FindClusters(pbmc, resolution = 0.5)
pbmc <- FindClusters(pbmc, resolution = 0.8)
pbmc <- FindClusters(pbmc, resolution = 1.2)
```

Each call adds a new column to the metadata named `RNA_snn_res.X` (where X is
the resolution value). We can visualize how cells move between clusters at
different resolutions using **clustree**:

```{r clustree, eval=FALSE, fig.alt="Clustree diagram showing how clusters split as resolution increases from 0.2 to 1.2. At low resolution a few large clusters exist, which progressively split into more subclusters at higher resolutions. Stable splits show clean branches while unstable splits show cells moving between multiple clusters."}
clustree(pbmc, prefix = "RNA_snn_res.")
```

In the clustree plot, each row is a resolution and each node is a cluster.
Arrows show how cells move between clusters as resolution increases. Look for:

- **Clean splits**: one cluster at a lower resolution cleanly divides into two
  at a higher resolution. This suggests a real biological distinction.
- **Unstable splits**: cells from one cluster scatter across multiple clusters
  at a higher resolution, or cells bounce between clusters. This suggests
  over-splitting.

Set the active cluster identity to our chosen resolution:

```{r set-resolution, eval=FALSE}
Idents(pbmc) <- "RNA_snn_res.0.5"
```

Save the clustered object for the next episode:

```{r save, eval=FALSE}
saveRDS(pbmc, file = "pbmc_clustered.rds")
```


## Exploring Clusters

Before formal cell type annotation (next episode), we can already preview cell
identity by overlaying known marker genes on the UMAP. If our clustering is
good, each marker should light up in a distinct cluster.

```{r feature-plot, eval=FALSE, fig.alt="Four UMAP plots showing expression of CD3D, MS4A1, LYZ, and GNLY. Each gene is highlighted in a different cluster region, indicating T cells, B cells, monocytes, and NK cells respectively."}
FeaturePlot(pbmc,
            features = c("CD3D", "MS4A1", "LYZ", "GNLY"),
            ncol = 2)
```

| Gene | Cell type | Expected pattern |
|------|:----------|:----------------|
| `CD3D` | T cells | Lights up the large T cell clusters (the biggest groups on the UMAP) |
| `MS4A1` | B cells | Lights up a compact cluster separate from T cells |
| `LYZ` | CD14+ Monocytes | Lights up monocyte clusters, distinct from lymphocytes |
| `GNLY` | NK cells | Lights up a cluster adjacent to but separate from T cells |

We can also view these markers as violin plots to see the expression
distribution across clusters:

```{r vln-markers, eval=FALSE, fig.alt="Violin plots showing expression of CD3D, MS4A1, LYZ, and GNLY across all clusters. Each marker shows high expression in one or two specific clusters and low or zero expression in the others."}
VlnPlot(pbmc,
        features = c("CD3D", "MS4A1", "LYZ", "GNLY"),
        ncol = 2,
        pt.size = 0)
```

Even without formal annotation, you can see that these markers segregate
cleanly into distinct clusters. CD3D marks several clusters (T cell subtypes),
MS4A1 is specific to one cluster (B cells), LYZ is strong in one or two
clusters (monocytes), and GNLY is specific to another (NK cells). This gives
us confidence that the clustering is capturing biologically meaningful
groupings. In the next episode, we will systematically identify marker genes
for every cluster and assign cell type labels.


::::::::::::::::::::::::::::::::::::: challenge

## Challenge 1: Choosing a Resolution with Clustree

Run clustering at resolutions 0.2, 0.5, 0.8, and 1.2 (we already did this
above). Use the `clustree` plot to answer:

1. At which resolution do clusters start splitting unstably?
2. Which resolution gives a number of clusters that best matches the known
   PBMC cell types (~8--10 major types)?
3. How many clusters do you get at each resolution?

```{r challenge1, eval=FALSE}
# Count clusters at each resolution
cat("Resolution 0.2:", length(unique(pbmc$RNA_snn_res.0.2)), "clusters\n")
cat("Resolution 0.5:", length(unique(pbmc$RNA_snn_res.0.5)), "clusters\n")
cat("Resolution 0.8:", length(unique(pbmc$RNA_snn_res.0.8)), "clusters\n")
cat("Resolution 1.2:", length(unique(pbmc$RNA_snn_res.1.2)), "clusters\n")

# Visualize side by side
p1 <- DimPlot(pbmc, group.by = "RNA_snn_res.0.2", label = TRUE) +
    NoLegend() + ggtitle("Resolution 0.2")
p2 <- DimPlot(pbmc, group.by = "RNA_snn_res.0.5", label = TRUE) +
    NoLegend() + ggtitle("Resolution 0.5")
p3 <- DimPlot(pbmc, group.by = "RNA_snn_res.0.8", label = TRUE) +
    NoLegend() + ggtitle("Resolution 0.8")
p4 <- DimPlot(pbmc, group.by = "RNA_snn_res.1.2", label = TRUE) +
    NoLegend() + ggtitle("Resolution 1.2")

(p1 + p2) / (p3 + p4)
```

::::::::::::::::::::::::: solution

Expected cluster counts:

| Resolution | Approximate clusters |
|:----------:|:--------------------:|
| 0.2 | 6--7 |
| 0.5 | 9--11 |
| 0.8 | 12--14 |
| 1.2 | 15--18 |

**Resolution 0.5** gives approximately 9--11 clusters, which aligns well with
the 8 major PBMC cell types (CD4+ T, CD8+ T, B cells, CD14+ monocytes,
FCGR3A+ monocytes, NK cells, dendritic cells, platelets). Some cell types may
be split into sub-clusters (e.g., naive vs. memory T cells), which is
biologically reasonable.

**Resolution 0.2** under-clusters: it merges CD4+ and CD8+ T cells, or
combines monocyte subtypes. You lose meaningful biological distinctions.

**Resolution 0.8** starts to over-split: the clustree plot shows T cell
clusters from resolution 0.5 splitting into sub-clusters, with some cells
moving between clusters in ways that are not stable. This does not mean 0.8 is
wrong -- it may reveal real sub-populations -- but the splits become harder to
annotate confidently.

**Resolution 1.2** clearly over-splits: you get 15+ clusters, many of which
are fragments of the same cell type. The clustree plot shows extensive
cross-cluster mixing at this resolution.

For this workshop we use **resolution 0.5** as a good balance between
capturing distinct cell types and avoiding excessive fragmentation.

:::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: challenge

## Challenge 2: Mapping Markers to Clusters

Use `FeaturePlot()` to visualize three markers on the UMAP:

- **CD3D** (T cells)
- **MS4A1** (B cells)
- **LYZ** (monocytes)

Then use the cluster-labeled UMAP to determine which cluster number(s)
correspond to each cell type.

```{r challenge2, eval=FALSE}
FeaturePlot(pbmc, features = c("CD3D", "MS4A1", "LYZ"), ncol = 3)
DimPlot(pbmc, reduction = "umap", label = TRUE, label.size = 5) + NoLegend()
```

:::::::::::::::::::::::: solution

By comparing the FeaturePlot panels with the labeled UMAP, you should see the
following pattern (exact cluster numbers may differ depending on your run):

| Marker | Cell type | Cluster(s) |
|--------|:----------|:-----------|
| CD3D | T cells | Clusters 0, 2, 4 (the largest groups on the UMAP). CD3D is expressed in both CD4+ and CD8+ T cells, so it marks multiple clusters. |
| MS4A1 | B cells | Cluster 3 (a compact, clearly separated group). MS4A1 (also known as CD20) is highly specific to B cells. |
| LYZ | CD14+ Monocytes | Cluster 1 (a large group on the opposite side of the UMAP from T cells). You may also see weaker LYZ expression in a smaller monocyte cluster (FCGR3A+). |

The fact that each marker cleanly maps to specific clusters confirms that
our clustering at resolution 0.5 is capturing meaningful biological groupings.
Notice that CD3D marks multiple clusters -- this is expected because there are
several T cell subtypes (CD4+ naive, CD4+ memory, CD8+) that form separate
clusters but all express the pan-T cell marker CD3D.

In the next episode, we will use `FindAllMarkers()` to systematically identify
the genes that define each cluster and assign proper cell type labels.

:::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: keypoints

- PCA reduces ~2,000 variable genes to ~15 principal components that capture the major axes of biological variation
- The elbow plot helps choose how many PCs to retain; for PBMCs, 15 PCs is typically sufficient
- UMAP provides an intuitive 2D visualization but should not be used to infer distances, cluster sizes, or quantitative relationships
- Graph-based clustering builds a shared nearest neighbor graph in PCA space and detects communities using the Louvain or Leiden algorithm
- The resolution parameter controls clustering granularity; clustree helps identify the resolution where clusters begin to split unstably

::::::::::::::::::::::::::::::::::::::::::::::::
