---
source: Rmd
title: 'Normalization and Feature Selection'
teaching: 40
exercises: 15
---

:::::::::::::::::::::::::::::::::::::: questions

- Why do we need to normalize scRNA-seq data before comparing cells?
- How does log normalization work and where are the results stored?
- What are highly variable features and why do we select them?
- What does scaling do and when should we regress out confounders?
- How does SCTransform differ from the LogNormalize workflow?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Apply log normalization to correct for differences in sequencing depth between cells
- Identify highly variable features using variance-stabilizing transformation
- Scale the data to prepare for PCA
- Describe SCTransform as an alternative to the three-step LogNormalize workflow

::::::::::::::::::::::::::::::::::::::::::::::::

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(Seurat)
library(ggplot2)
library(patchwork)
```

## Loading the Filtered Data

We start from the filtered Seurat object saved at the end of the previous
episode. This object contains only cells that passed our QC thresholds.

```{r load-data, eval=FALSE}
pbmc <- readRDS("pbmc_filtered.rds")
pbmc
```

:::::::::::::::::::::::::::::::::::::::::: spoiler

## Expected output

```output
An object of class Seurat
23694 features across 10000 samples within 1 assay
Active assay: RNA (23694 features, 0 variable features)
 1 layer present: counts
```

The object has approximately 23,700 genes and 10,000 cells. Only the `counts`
layer is populated at this stage. By the end of this episode we will fill in
the `data` and `scale.data` layers as well.

::::::::::::::::::::::::::::::::::::::::::::::::::

Right now, the only expression data we have is raw UMI counts. Before we can
compare gene expression between cells, we need to address a fundamental
problem: **cells are sequenced to different depths**.

Consider two cells that are biologically identical -- they express the same
genes at the same levels. If Cell A was sequenced to 5,000 total UMIs and Cell
B to 10,000 total UMIs, every gene in Cell B will appear to have roughly twice
the counts of Cell A. This is purely a technical artifact of sequencing depth,
not a biological difference. Normalization corrects for this so that expression
values are comparable across cells.


## Log Normalization

The standard normalization in Seurat is **LogNormalize**. It applies a simple
three-step transformation to each cell independently:

1. **Divide** each gene's count by the cell's total UMI count
2. **Multiply** by a scale factor (default 10,000) so the values are not
   tiny fractions
3. **Log-transform** with `log1p` (natural log of 1 + x) to compress the
   dynamic range

The formula for a single gene *g* in cell *c* is:

**normalized(g, c) = log(1 + count(g, c) / total_counts(c) x 10000)**

```{r normalize, eval=FALSE}
pbmc <- NormalizeData(pbmc,
                      normalization.method = "LogNormalize",
                      scale.factor = 10000)
```

| Parameter | Value | Description |
|-----------|-------|-------------|
| `normalization.method` | `"LogNormalize"` | Divide by total counts, multiply by scale factor, log1p transform. |
| `scale.factor` | `10000` | Multiplicative factor applied after dividing by total counts. The default 10,000 means normalized values are in "counts per 10,000" before log transformation. |

After normalization, the results are stored in the `data` layer of the RNA
assay:

```{r check-layers, eval=FALSE}
pbmc
```

:::::::::::::::::::::::::::::::::::::::::: spoiler

## Expected output

```output
An object of class Seurat
23694 features across 10000 samples within 1 assay
Active assay: RNA (23694 features, 0 variable features)
 2 layers present: counts, data
```

Notice that we now have **2 layers**: `counts` (raw) and `data` (normalized).

::::::::::::::::::::::::::::::::::::::::::::::::::

You can access each layer directly:

```{r access-layers, eval=FALSE}
# Raw counts for the first 5 genes in the first 3 cells
pbmc[["RNA"]]$counts[1:5, 1:3]

# Normalized values for the same genes and cells
pbmc[["RNA"]]$data[1:5, 1:3]
```

### Visualizing the effect of normalization

To see what normalization does, let's compare the distribution of a gene before
and after normalization. We'll look at **LYZ**, a marker for CD14+ monocytes
that has high and variable expression across cells.

```{r norm-comparison, eval=FALSE, fig.alt="Two histograms side by side comparing raw UMI counts and log-normalized expression for the LYZ gene. Raw counts are right-skewed with a spike at zero. Normalized values are more spread out and approximately normal for expressing cells."}
# Extract raw and normalized values for LYZ
lyz_raw <- pbmc[["RNA"]]$counts["LYZ", ]
lyz_norm <- pbmc[["RNA"]]$data["LYZ", ]

p1 <- ggplot(data.frame(x = as.numeric(lyz_raw)), aes(x = x)) +
    geom_histogram(bins = 50, fill = "steelblue", color = "white") +
    labs(title = "LYZ - Raw counts", x = "UMI counts", y = "Number of cells") +
    theme_minimal()

p2 <- ggplot(data.frame(x = as.numeric(lyz_norm)), aes(x = x)) +
    geom_histogram(bins = 50, fill = "darkred", color = "white") +
    labs(title = "LYZ - Log-normalized", x = "Normalized expression", y = "Number of cells") +
    theme_minimal()

p1 + p2
```

The raw count distribution is heavily right-skewed with a large spike at zero
(cells that don't express LYZ -- mostly non-monocytes). After log
normalization, the non-zero values are spread more evenly and the influence
of sequencing depth differences is reduced.


## Identifying Variable Features

The human genome contains roughly 20,000 protein-coding genes, but not all of
them are informative for distinguishing cell types. Many genes are:

- **Housekeeping genes** expressed at similar levels in every cell (e.g., ACTB,
  GAPDH). These carry no information about cell identity.
- **Low-expression genes** detected in only a handful of cells. These are too
  noisy to contribute meaningfully.
- **Constant genes** that show little variation across cells.

**Highly variable features (HVGs)** are genes that vary substantially from cell
to cell. These are the genes most likely to capture biological differences
between cell types. By selecting the top 2,000 variable features, we focus
downstream analysis (PCA, clustering) on the genes that matter most, while
reducing noise and computation time.

```{r find-variable, eval=FALSE}
pbmc <- FindVariableFeatures(pbmc,
                             selection.method = "vst",
                             nfeatures = 2000)
```

| Parameter | Value | Description |
|-----------|-------|-------------|
| `selection.method` | `"vst"` | Variance-stabilizing transformation. Fits a mean-variance relationship across genes using local regression, then selects genes with the highest standardized variance (i.e., genes that vary more than expected given their expression level). |
| `nfeatures` | `2000` | Number of variable features to select. 2,000 is the standard default for most scRNA-seq analyses. |

Let's see which genes were selected:

```{r top-variable, eval=FALSE}
head(VariableFeatures(pbmc), 10)
```

:::::::::::::::::::::::::::::::::::::::::: spoiler

## Expected output

```output
 [1] "S100A9" "LYZ"    "S100A8" "GNLY"   "FTL"    "NKG7"   "IGKC"   "FTH1"
 [9] "CST3"   "TYROBP"
```

The top variable features include known immune cell markers: S100A9 and LYZ
(monocytes), GNLY and NKG7 (NK cells), IGKC (B cells), and CST3 (dendritic
cells). This makes sense -- these genes are highly expressed in specific cell
types but absent in others, producing high cell-to-cell variance.

::::::::::::::::::::::::::::::::::::::::::::::::::

Visualize the variable feature selection with a mean-variance plot:

```{r variable-plot, eval=FALSE, fig.alt="Scatter plot of mean expression versus standardized variance for all genes. The 2000 selected variable features are shown in red, with the top 10 labeled. Non-variable genes cluster along the bottom."}
vf_plot <- VariableFeaturePlot(pbmc)
top10 <- head(VariableFeatures(pbmc), 10)
LabelPoints(plot = vf_plot, points = top10, repel = TRUE)
```

In this plot, each point is a gene. The x-axis shows mean expression across
cells and the y-axis shows standardized variance. The red points are the 2,000
selected variable features. Genes in the upper right are both highly expressed
and highly variable -- these are strong cell-type markers. Genes along the
bottom have low variance relative to their expression level and are excluded.


## Scaling

The final step before PCA is **scaling**. `ScaleData()` applies a z-score
transformation to each gene across all cells: it subtracts the mean expression
and divides by the standard deviation. After scaling, each gene has mean 0 and
standard deviation 1.

Why is this necessary? Without scaling, PCA would be dominated by highly
expressed genes simply because they have larger absolute values, not because
they carry more biological information. Scaling puts all genes on equal footing
so that PCA identifies the axes of true biological variation rather than
expression magnitude.

```{r scale, eval=FALSE}
pbmc <- ScaleData(pbmc)
```

By default, `ScaleData()` only scales the variable features (the 2,000 genes
selected above), not all ~23,000 genes. This is faster and sufficient for PCA,
which only uses variable features. If you later need scaled values for all
genes (e.g., for a heatmap of a non-variable gene), you can re-run
`ScaleData()` with `features = rownames(pbmc)`.

After scaling, a third layer appears in the object:

```{r check-scale, eval=FALSE}
pbmc
```

:::::::::::::::::::::::::::::::::::::::::: spoiler

## Expected output

```output
An object of class Seurat
23694 features across 10000 samples within 1 assay
Active assay: RNA (23694 features, 2000 variable features)
 3 layers present: counts, data, scale.data
```

All three layers are now populated: `counts` (raw), `data` (log-normalized),
and `scale.data` (z-scored variable features).

::::::::::::::::::::::::::::::::::::::::::::::::::

### Regressing out confounders

`ScaleData()` can optionally **regress out** unwanted sources of variation
using the `vars.to.regress` parameter. A common use case is regressing out
mitochondrial percentage so that cell damage does not influence clustering:

```{r scale-regress, eval=FALSE}
# Optional: regress out mitochondrial percentage
pbmc_regressed <- ScaleData(pbmc, vars.to.regress = "percent.mt")
```

This fits a linear model for each gene with `percent.mt` as a covariate and
uses the residuals as the scaled expression values. The effect is that
variation driven by mitochondrial content is removed before PCA.

**When to regress:**

- Mitochondrial percentage is driving a major axis of variation in PCA (i.e.,
  one of the top PCs strongly correlates with percent.mt)
- You see cells grouping by mitochondrial content rather than cell type

**When NOT to regress:**

- Mitochondrial content correlates with real biology. Some cell types
  (e.g., cardiomyocytes, metabolically active immune cells) naturally have
  higher mitochondrial gene expression. Regressing it out would remove
  legitimate biological signal.
- The QC filtering in the previous episode already removed high-mt cells, and
  the remaining variation is modest.

For this workshop, we proceed **without** regression because our QC filtering
already removed cells with `percent.mt > 15`, and the remaining variation does
not dominate the PCA.

Save the normalized object for the next episode:

```{r save, eval=FALSE}
saveRDS(pbmc, file = "pbmc_normalized.rds")
```


## SCTransform as an Alternative

:::::::::::::::::::::::::::::::::::::::::: callout

## SCTransform: a one-step alternative

The three-step LogNormalize workflow (`NormalizeData` + `FindVariableFeatures` +
`ScaleData`) works well for most datasets, but it has limitations. The log
transformation can over-stabilize variance for lowly expressed genes and
under-stabilize it for highly expressed genes. This matters most when there
is high technical noise in the data.

**SCTransform** addresses this by using **regularized negative binomial
regression**. It models each gene's expression as a function of total UMI
counts per cell, estimates the expected variance at each expression level, and
returns Pearson residuals that serve as normalized, variance-stabilized values.
A single function call replaces all three steps:

```r
pbmc_sct <- SCTransform(pbmc, vst.flavor = "v2")
```

| Parameter | Value | Description |
|-----------|-------|-------------|
| `vst.flavor` | `"v2"` | Use the improved v2 algorithm (Choudhary & Satija, 2022) with better variance estimation for sparse data. |

Key differences from LogNormalize:

- **Replaces three steps** with one: normalization, feature selection, and
  scaling are all handled internally.
- **Selects 3,000 variable features** by default (vs. 2,000 for
  `FindVariableFeatures`), which can improve resolution for complex tissues.
- **Better variance stabilization** across the full range of expression levels.
- **Stored in a separate assay**: results go into an assay called `SCT` rather
  than modifying the `RNA` assay. The `RNA` assay retains the raw counts for
  reference.

**When to use SCTransform:**

- For publication-quality analyses, especially with complex tissues or datasets
  with high technical variance.
- When you observe that LogNormalize produces clusters driven by sequencing
  depth rather than biology.

**When LogNormalize is sufficient:**

- For exploratory analysis and quick overviews.
- When the dataset is well-behaved and you want a simpler, faster workflow.
- For direct comparability with older analyses and tutorials.

In this workshop, we use **LogNormalize** throughout for simplicity and because
it produces excellent results with this well-characterized PBMC dataset.
However, SCTransform is the preferred method for most production analyses.

:::::::::::::::::::::::::::::::::::::::::::::::::::


::::::::::::::::::::::::::::::::::::: challenge

## Challenge 1: LogNormalize vs. SCTransform Variable Features

Run both normalization approaches on the same filtered dataset and compare the
top 20 variable features from each method. How many genes appear in both lists?

```{r challenge1, eval=FALSE}
# LogNormalize pathway (already done above)
lognorm_top20 <- head(VariableFeatures(pbmc), 20)

# SCTransform pathway
pbmc_sct <- SCTransform(pbmc, vst.flavor = "v2")
sct_top20 <- head(VariableFeatures(pbmc_sct), 20)

# Compare
cat("LogNormalize top 20:\n")
print(lognorm_top20)
cat("\nSCTransform top 20:\n")
print(sct_top20)
cat("\nGenes in both lists:\n")
shared <- intersect(lognorm_top20, sct_top20)
print(shared)
cat("\nOverlap:", length(shared), "out of 20\n")
```

:::::::::::::::::::::::: solution

You should find that approximately **12--14 of the top 20 genes overlap**
(~60--70%), which means both methods broadly agree on which genes are most
variable. The shared genes typically include strong cell-type markers like
S100A9, LYZ, GNLY, NKG7, and IGKC.

The differences arise because the two methods define "variable" differently:

- **vst** (LogNormalize) fits a mean-variance curve and selects genes with the
  highest residual variance. It can be influenced by a few outlier cells with
  extreme counts.
- **SCTransform** uses a negative binomial model that is more robust to
  outliers and better handles the mean-variance relationship at low expression
  levels.

Despite these differences in the top-ranked features, the downstream results
(PCA, clustering, cell type identification) are usually very similar because
both methods capture the major axes of biological variation. The differences
tend to matter more for subtle distinctions between closely related cell states.

:::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: challenge

## Challenge 2: Housekeeping vs. Variable Gene Behavior

Plot the expression of **ACTB** (a housekeeping gene) and **LYZ** (a variable
monocyte marker) before and after normalization. What changes for each gene,
and why?

```{r challenge2, eval=FALSE}
# Extract raw and normalized values
actb_raw  <- as.numeric(pbmc[["RNA"]]$counts["ACTB", ])
actb_norm <- as.numeric(pbmc[["RNA"]]$data["ACTB", ])
lyz_raw   <- as.numeric(pbmc[["RNA"]]$counts["LYZ", ])
lyz_norm  <- as.numeric(pbmc[["RNA"]]$data["LYZ", ])

p1 <- ggplot(data.frame(x = actb_raw), aes(x)) +
    geom_histogram(bins = 50, fill = "steelblue") +
    labs(title = "ACTB - Raw counts") + theme_minimal()
p2 <- ggplot(data.frame(x = actb_norm), aes(x)) +
    geom_histogram(bins = 50, fill = "darkred") +
    labs(title = "ACTB - Normalized") + theme_minimal()
p3 <- ggplot(data.frame(x = lyz_raw), aes(x)) +
    geom_histogram(bins = 50, fill = "steelblue") +
    labs(title = "LYZ - Raw counts") + theme_minimal()
p4 <- ggplot(data.frame(x = lyz_norm), aes(x)) +
    geom_histogram(bins = 50, fill = "darkred") +
    labs(title = "LYZ - Normalized") + theme_minimal()

(p1 + p2) / (p3 + p4)
```

:::::::::::::::::::::::: solution

**ACTB (housekeeping gene):**

- **Before normalization:** the raw count distribution is right-skewed because
  cells with higher sequencing depth have proportionally more ACTB counts. The
  apparent variation is mostly technical (sequencing depth), not biological.
- **After normalization:** the distribution becomes tighter and more uniform.
  Because ACTB is expressed at similar levels in all cell types, once you
  correct for sequencing depth the remaining variation is small. This is exactly
  why ACTB is **not** selected as a variable feature -- it doesn't help
  distinguish cell types.

**LYZ (variable gene):**

- **Before normalization:** there is a large spike at zero (non-monocyte cells
  that don't express LYZ) and a spread of non-zero values. Part of the
  variation in non-zero values is technical (depth) and part is biological
  (different amounts of LYZ in monocytes).
- **After normalization:** the zero spike remains (you can't normalize away
  true absence of expression), but the non-zero values are corrected for depth.
  The biological variation between monocytes and non-monocytes is preserved and
  stands out more clearly. This is why LYZ **is** selected as a highly variable
  feature.

The key takeaway: normalization makes expression comparable across cells by
removing sequencing depth effects. Housekeeping genes become more uniform
(confirming they don't vary biologically), while cell-type marker genes retain
their meaningful variation.

:::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: keypoints

- Log normalization corrects for sequencing depth by dividing by total counts, scaling, and log-transforming each cell independently
- Highly variable features are the ~2,000 genes with the most cell-to-cell variation, capturing biological differences while excluding housekeeping genes and noise
- Scaling (z-score transformation) centers and standardizes gene expression so that PCA is not dominated by highly expressed genes
- Regressing out confounders like percent.mt during scaling is optional and should only be done when technical variation is obscuring biology
- SCTransform is a one-step alternative that uses regularized negative binomial regression and is recommended for publication-quality analyses

::::::::::::::::::::::::::::::::::::::::::::::::
